Future Cosmic Microwave Background delensing with galaxy surveys.

### 2017-07-10 Notes/galaxy_delensing.tex
S4 resu

The cosmic microwave background (CMB) polarization is a promising experimental dataset to test the inflationary paradigm and to probe the physics of the early universe. 
A particular component, the so-called B-modes, is indeed a direct signature of a prediction of inflation: the presence of gravitational waves in the early universe. However, reducing the instrumental noise in future experiments will not be enough to detect this signal. 
Secondary effects in the low redshift universe will also produce non-primordial B-modes adding confusion to the inflationary signal. In particular, the gravitational interactions of CMB photons with large scale structures will distort the primordial E-modes adding a lensing B-mode component to the primordial signal.
Removing the lensing part ('delensing') from the measurement of CMB B-modes will then be necessary to constrain the amplitude of the primordial gravitational waves.
Here we discuss the role of current and future large scale structure surveys in a multi-tracers approach to delensing that will improve the reconstruction of the lensing potential that lenses the CMB photons and, as a consequence, the delensing efficiency. 
We quantify this by the improvement due to delensing on the constraints on the inflationary tensor perturbations amplitude and shape ($r$ and $n_{t}$). We find that, in general, galaxy surveys should be splitted into tomographic bins as this can improve the correlation with CMB lensing by 30\%. 
Considering currently available survey, a DES-like galaxy survey can remove about $14\%$ of the lensing signal. The CMB experiments of the current generation will particularly benefit from large scale structure tracers that, once properly combined, will have a better performance than a CMB internal reconstruction.
With the decrease of the instrumental noise, the CMB internal reconstruction will increase its efficiency and the fraction of removed lensing B-modes with CMB alone will rapidly improve from the current level of Planck ($8\%$) and SPTPol ($35\%$) to 3G (56\%) and CMB S4 (85\%) level.  
Nevertheless optical galaxy surveys will still play an important role even for CMB S4. In particular LSST can a achieve a delensing performance comparable to a 3G CMB experiment but and completely different systematics. This will be important to prove the robustness against systematics of an eventual detection of primordial B-modes.
% maybe non linear scales?
\end{abstract}

@@

\reftab{S4} shows the improvement on the inflationary constraint given the cross-correlation factors described above. 
LSST alone will be able to remove almost half of the B-mode lensing power. This will lead to an improvement in $\sigma(r)$ of a factor of 1.8. If we add all the other infrared and optical LSS tracers previously included (i.e. excluding SKA) the improvement factor increase to 2.3.
These gives an important opportunity: for sufficiently high value of $r$ it will be possible to confirm a detection of primordial gravity waves using spectra delensed with LSS, completely independent of CMB internal data. 
Similarly to \cite{namikawa:2016a} we find that SKA phase-1 radio survey will remove 53\% of the lensing power and, once combined with optical and infrared surveys, this will allow the level of power removed using LSS (62\%) to surpass the level of CMB S3 internal reconstruction.
Finally, as previously mentioned, CMB S4 will be the main source of delensing with the ability of removing up to more than 80\% of the B-modes power.

\relook{FINISH}

\begin{table}
\caption{$\alpha$: Gen 4 experiments}
  \begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
LSST (48\%) &1.88 & 1.81  &  \\ \hline
LSS  (60\%) &2.47 & 2.31  & \\ \hline
SKA (10$\mu$Jy) (53\%) &2.18& 2.06  & \\ \hline
SKA (0.1$\mu$Jy) (77\%) &4.6& 3.96  & \\ \hline
LSS + SKA (10$\mu$Jy)  (62\%) &2.74 & 2.53  &  \\ \hline
LSS + SKA (0.1$\mu$Jy)  (80\%) &4.66 & 3.96  &  \\ \hline
CMB S4 (81\%) & 4.63 & 3.96  &  \\ \hline
LSS+CMB (69\%)& 2.8 & 1.8&  \\ \hline
\end{tabular}
\label{tab:S4}
\end{table}
%=======================================
\begin{table}
\caption{$\alpha$: Gen 4 experiments}
  \begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
LSST (10\%) &1.12 & 1.08  & 1.03 \\ \hline
SKA (0.1$\mu$Jy) (40\%) &1.73& 1.42  & 1.2\\ \hline
CMB S4 (56\%) &2& 1.57  & 1.21 \\ \hline
LSS+CMB (69\%)& 2.8 & 1.8& 1.31  \\ \hline
\end{tabular}
\label{tab:S4_2}
\end{table}
%=======================================
%\subsection{Bias uncertainties degradation}
%The uncertainties in the theoretical assumptions used to model the galaxies can cause a degradation of the improvement of the inflationary constraint of delensed spectra.
%In this section, we quantify this effect.
%We will now marginalize over unknown galaxies parameters, but we will use a full dataset of CMB and galaxies data. The idea is that as shown in the low level of noise in galaxy surveys might allow us to calibrate them internally.
%We will use a Fisher approach the Fisher matrix is:
%\beqn
%F_{pq} &=& \sum_{l_{a} =l^{BB}_{\mathrm{min}}}^{l^{BB}_{\mathrm{max}}}  \sum_{l_{b} =l^{BB}_{\mathrm{min}}}^{l^{BB}_{\mathrm{max}}}   \frac{\partial {C}_{l_a}^{BB,\mathrm{del}}}{\partial \theta_p } \left[ \mathrm{Cov}^{BB,BB}\right]_{l_a, l_b} ^{-1} \frac{\partial {C}_{l_b}^{BB,\mathrm{del}}}{\partial \theta_q }\nonumber \\&+&  \sum_j \frac{\frac{\partial {C}_j^{\kappa I}}{\partial \theta_p } \frac{\partial {C}_j^{\kappa I}}{\partial \theta_q }}{(\Delta C_j^{\kappa I})^2}
%+\sum_j \frac{\frac{\partial {C}_j^{I I}}{\partial \theta_p } \frac{\partial {C}_j^{I I}}{\partial \theta_q }}{(\Delta C_j^{I  I})^2}
%\eeqn
%
%\beq
%\alpha_{\mathrm{marginalized}} = \sigma_0(r) / \sigma_{\mathrm{marginalized/delensed}}(r)
%\eeq
%
%
%where the parameter array contains both the tensor to scalar rate $\theta=r$, and the galaxy surveys parameters like the bias $b_i$ or $p_i$ \footnote{In our analysis, since the fiducial value of $r$ is zero,
%the derivative of $C_\ell^{\rm BB,res}$ is non-zero if $\theta_i=r$.}.
%
%We compute the derivatives of the power spectra as described
\section{Conclusions}
\label{sec:concl}
The ability to separate the lensing component of the CMB B-modes from a possible primordial inflationary signal ("delensing") is necessary to test the inflationary paradigm using the next generation of CMB polarization experiments. 
To do so, we need to accurately reconstruct the large scale structures that lens the CMB in order to build a template of the lensing B-modes.
In this paper, we studied the potential impact of large-scale structure galaxy surveys in this important endeavor.
%For current experiment, the cosmic infrared background had already proven to be very efficient in delensing \cite{larsen:2016,manzotti:2017}.

For ongoing experiments, we find that combining LSS tracers with CMB internal reconstruction will be particularly beneficial.
An optical survey like DES alone is able to remove $14\%$ of the lensing component and together with WISE($8\%$) and the CIB ($27\%$) will allow to remove $36\%$ of the contaminant signal using only LSS survey.
Depending on the CMB noise and the amount of galactic foreground cleaning these can correspond to an improvement of 30\% in the constrain on the tensor to scalar ratio $r$.
In the future, the decreasing level of instrumental noise of CMB experiments will dramatically improve the internal reconstruction of the structures lensing the CMB. 
The fraction of removed lensing B-modes will rapidly improve from the current level of Planck ($8\%$) and SPTPol ($35\%$) to 3G (56\%) and CMB S4 (85\%) level. 
Already in the 3G era, the CMB internal reconstruction will be the main source of delensing. However it will still be less efficient than galaxies at small scales$\ell>500$. For this reason, combining galaxy survey with the CMB will push the fraction of removed power from $57\%$ to $68\%$.

Even for CMB S4, galaxy surveys will still play an important role.
For example, a tomographically binned LSST-survey will remove 56\% of the lensing power. While this performance is comparable to a Stage-3 CMB it will allow to probe the robustness against systematics of an eventual detection.
Indeed delensing with just CMB data will require a careful study of possible biases and systematics effects due to the use of the same source (the CMB) we are delensing to reconstruct the lensing effect itself \cite{carron:2017,sehgal:2017,namikawa:2017}.
For this reason, efficient galaxies tracers are not only useful in the short term but will also play a role in cross-checking these internal biases.

Given the high level of correlation of future galaxy surveys and CMB lensing, it is also worth looking for potential application besides the detection of inflationary B-modes.
For example, LSS delensing allows to selectively remove from the CMB  the gravitational effect coming from the non-linear structures in the low-redshift universe. 
Removing this component will reduce the level of non-linearities both in the CMB and in the CMB lensing power spectrum providing an alternative route to exploit all the measurable modes without having to model the non-linear component at small angular scales.

%=================================================================
\section{Acknowledgments}
We thank S. Passaglia, S. Dodelson, K. Story and K. Wu for useful discussion and their review of the draft. 
Furthermore we thank D. Alonso for his feedback on radio continuum surveys.

>>>>

The cosmic microwave background (CMB) polarization is a promising experimental dataset to test the inflationary paradigm and to probe the physics of the early universe. 
A particular component, the so-called B-modes, is indeed a direct signature of a prediction of inflation: the presence of gravitational waves in the early universe. However, reducing the instrumental noise in future experiments will not be enough to detect this signal. 
Secondary effects in the low redshift universe will also produce non-primordial B-modes adding confusion to the inflationary signal. In particular, the gravitational interactions of CMB photons with large scale structures will distort the primordial E-modes adding a lensing B-mode component to the primordial signal.
Removing the lensing part ('delensing') from the measurement of CMB B-modes will then be necessary to constrain the amplitude of the primordial gravitational waves.
Here we discuss the role of current and future large scale structure surveys in a multi-tracers approach to delensing that will improve the reconstruction of the lensing potential that lenses the CMB photons and, as a consequence, the delensing efficiency. 
We quantify this by the improvement due to delensing on the constraints on the inflationary tensor perturbations amplitude and shape ($r$ and $n_{t}$). We find that, in general, galaxy surveys should be splitted into tomographic bins as this can improve the correlation with CMB lensing by 30\%. 
Considering currently available survey, a DES-like galaxy survey can remove about $14\%$ of the lensing signal. The CMB experiments of the current generation will particularly benefit from large scale structure tracers that, once properly combined, will have a better performance than a CMB internal reconstruction.
With the decrease of the instrumental noise, the CMB internal reconstruction will increase its efficiency and the fraction of removed lensing B-modes with CMB alone will rapidly improve from the current level of Planck ($8\%$) and SPTPol ($35\%$) to 3G (56\%) and CMB S4 (85\%) level.  
Nevertheless optical galaxy surveys will still play an important role even for CMB S4. In particular an LSST-like survey can a achieve a delensing performance comparable to a 3G CMB experiment but with completely different systematics. This will be important to prove the robustness against systematics of an eventual detection of primordial B-modes.
% maybe non linear scales?
\end{abstract}

@@

\reftab{S4} shows the improvement on the inflationary constraint given the cross-correlation factors described above. 
LSST alone will be able to remove almost half of the B-mode lensing power. This will lead to an improvement in $\sigma(r)$ of a factor of 1.8. If we add all the other infrared and optical LSS tracers previously included (i.e. excluding SKA) the improvement factor increase to 2.3.
These gives an important opportunity: for sufficiently high value of $r$ it will be possible to confirm a detection of primordial gravity waves using spectra delensed with LSS, completely independent of CMB internal data. 
Similarly to \cite{namikawa:2016a} we find that SKA phase-1 radio survey will remove 53\% of the lensing power and, once combined with optical and infrared surveys, this will allow the level of power removed using LSS (62\%) to surpass the level of the CMB S3 internal reconstruction.
Finally, as previously mentioned, CMB S4 will be the main source of delensing with the ability of removing up to more than 80\% of the B-modes power. This will bring an almost factor of 4 improvement in the null test case even with realistic noise. Combining with LSS will only increase the removed power compared to CMB only by $\simeq 5\%$ but it will be highly beneficial to test the robustness of the final result.

We find the improvement in $\sigma(n_{t})$ to be very similar to the one in $\sigma(r)$ for all the CMB-S4 cases. This is reasonable because, given the level of noise in CMB-S4, the variance at angular scales $\ell<200$ are dominated by instrumental noise and not by the lensing component. Independent of the delensing level, the small value allowed for $r$ will not allow to test the consistency relations.

\begin{table}
\caption{$\alpha$: Gen 4 experiments}
\begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
LSST (48\%) &1.88 & 1.81  &  \\ \hline
LSS  (60\%) &2.47 & 2.31  & \\ \hline
SKA (10$\mu$Jy) (53\%) &2.18& 2.06  & \\ \hline
SKA (0.1$\mu$Jy) (77\%) &4.6& 3.96  & \\ \hline
LSS + SKA (10$\mu$Jy)  (62\%) &2.74 & 2.53  &  \\ \hline
LSS + SKA (0.1$\mu$Jy)  (80\%) &4.66 & 3.96  &  \\ \hline
CMB S4 (81\%) & 4.63 & 3.96  &  \\ \hline
LSS+CMB (86\%)& 5.2 &  &  \\ \hline
\end{tabular}
\label{tab:S4}
\end{table}
%=======================================
%\begin{table}
%\caption{$\alpha$: Gen 4 experiments}
%  \begin{tabular}{|c | c | c | c|}
%\hline
%Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
%LSST (10\%) &1.12 & 1.08  & 1.03 \\ \hline
%SKA (0.1$\mu$Jy) (40\%) &1.73& 1.42  & 1.2\\ \hline
%CMB S4 (56\%) &2& 1.57  & 1.21 \\ \hline
%LSS+CMB (69\%)& 2.8 & 1.8& 1.31  \\ \hline
%\end{tabular}
%\label{tab:S4_2}
%\end{table}
%=======================================
%\subsection{Bias uncertainties degradation}
%The uncertainties in the theoretical assumptions used to model the galaxies can cause a degradation of the improvement of the inflationary constraint of delensed spectra.
%In this section, we quantify this effect.
%We will now marginalize over unknown galaxies parameters, but we will use a full dataset of CMB and galaxies data. The idea is that as shown in the low level of noise in galaxy surveys might allow us to calibrate them internally.
%We will use a Fisher approach the Fisher matrix is:
%\beqn
%F_{pq} &=& \sum_{l_{a} =l^{BB}_{\mathrm{min}}}^{l^{BB}_{\mathrm{max}}}  \sum_{l_{b} =l^{BB}_{\mathrm{min}}}^{l^{BB}_{\mathrm{max}}}   \frac{\partial {C}_{l_a}^{BB,\mathrm{del}}}{\partial \theta_p } \left[ \mathrm{Cov}^{BB,BB}\right]_{l_a, l_b} ^{-1} \frac{\partial {C}_{l_b}^{BB,\mathrm{del}}}{\partial \theta_q }\nonumber \\&+&  \sum_j \frac{\frac{\partial {C}_j^{\kappa I}}{\partial \theta_p } \frac{\partial {C}_j^{\kappa I}}{\partial \theta_q }}{(\Delta C_j^{\kappa I})^2}
%+\sum_j \frac{\frac{\partial {C}_j^{I I}}{\partial \theta_p } \frac{\partial {C}_j^{I I}}{\partial \theta_q }}{(\Delta C_j^{I  I})^2}
%\eeqn
%
%\beq
%\alpha_{\mathrm{marginalized}} = \sigma_0(r) / \sigma_{\mathrm{marginalized/delensed}}(r)
%\eeq
%
%
%where the parameter array contains both the tensor to scalar rate $\theta=r$, and the galaxy surveys parameters like the bias $b_i$ or $p_i$ \footnote{In our analysis, since the fiducial value of $r$ is zero,
%the derivative of $C_\ell^{\rm BB,res}$ is non-zero if $\theta_i=r$.}.
%
%We compute the derivatives of the power spectra as described
\section{Conclusions}
\label{sec:concl}
The ability to separate the lensing component of the CMB B-modes from a possible primordial inflationary signal ("delensing") is necessary to test inflation using the next generation of CMB polarization experiments. 
To delens, we need to accurately reconstruct the large scale structures that lens the CMB in order to build a template of the expected lensing B-modes in the observed patch.
In this paper, we studied the potential impact of large-scale structure galaxy surveys in this important endeavor.
%For current experiment, the cosmic infrared background had already proven to be very efficient in delensing \cite{larsen:2016,manzotti:2017}.

For ongoing experiments, we find that combining LSS tracers with CMB internal reconstruction will be particularly beneficial.
An optical survey like DES alone is able to remove $14\%$ of the B-mode lensing power and together with WISE($8\%$) and the CIB ($27\%$) will allow to remove $36\%$ of the power using only LSS survey.
Depending on the CMB instrumental noise and the amount of galactic foreground cleaning, delensing using LSS survey can correspond to an improvement of 30\% in the constrain on the tensor to scalar ratio $r$.
In the future, the decreasing level of instrumental noise of CMB experiments will dramatically improve the internal reconstruction of the structures lensing the CMB. 
The fraction of removed lensing B-modes will rapidly improve from the current level of Planck ($8\%$) and SPTPol ($35\%$) to 3G (56\%) and CMB S4 (85\%) level. 
Already in the 3G era, the CMB internal reconstruction will be the main source of delensing. However it will still be less efficient than galaxies at small scales$\ell>500$. For this reason, combining galaxy survey with the CMB will push the fraction of removed power from $57\%$ to $68\%$.
Even for CMB S4, galaxy surveys will still play an important role.
For example, a tomographically binned LSST-survey will remove 56\% of the lensing power. While this performance is comparable to a Stage-3 CMB it will allow to probe the robustness against systematics of an eventual detection.
Indeed delensing with just CMB data will require a careful study of possible biases and systematics effects due to the use of the same source (the CMB) that we are trying to delens to reconstruct the lensing effect itself \cite{carron:2017,sehgal:2017,namikawa:2017}.
For this reason, efficient galaxies tracers are not only useful in the short term but will also play a role in cross-checking these internal biases in the future.

Given the high level of correlation of future galaxy surveys and CMB lensing, it is also worth looking for potential application besides the detection of inflationary B-modes.
For example, delensing with galaxy surveys allows to selectively remove from the CMB only the gravitational effect coming from the non-linear structures in the low-redshift universe. 
Removing this component will reduce the level of non-linearities both in the CMB and in the CMB lensing power spectrum providing an alternative route to exploit all the measurable modes without having to model the non-linear component at small angular scales.

%=================================================================
\section{Acknowledgments}
We thank S. Dodelson, S. Passaglia, K. Story and K. Wu for useful discussion and their review of the draft. 
Furthermore we thank D. Alonso for his feedback on radio continuum surveys.
